Many-body approaches to the homogeneous electron gas in two and three dimensions

### 2018-02-15 doc/ElectronGas/paper.tex
Added electron gas  paper

@@

Let us consider the initial Hamiltonian
\[
 \hat{H} = \hat{H}^{\rm d} + \Ho,
\]
where $\hat{H}^{\rm d}$ and $\Ho$ denote its ``diagonal'' and
``off-diagonal'' parts, respectively.
 Introducing a flow parameter $s$, there exits a unitary
 transformation $U_s$, such that
\begin{equation}
 \hat{H}_s = U_s^\dagger \hat{H} U_s \equiv \hat{H}^{\rm d}_s + \Ho_s,
\end{equation}
with the relations $U_{s=0} = \mathbf{1}$, and $\hat{H}_{s= 0} =\hat{H}$.  
The transformation $U_s$ is parametrized as
\[
U_s = T_s \exp \left(\int_0^s \! ds'\hat{\eta}_{s'} \right),
\]
where the anti-hermitian operator $\hat{\eta}_s$ serves as generator
of the transformation. With $T_s$ we denote $s$-ordering, which is
defined equivalently to usual time-ordering.  Taking the derivative of
$\hat{H}_s$ with respect to $s$ gives
\begin{equation}
 \frac{d \hat{H}_s}{ds} = \frac{d U_s}{ds}\hat{H} U_s^\dagger + U_s
 \hat{H} \frac{d U_s^\dagger}{ds}.
\label{eq:flow_long}
\end{equation}
Utilizing that for our particular form of $U_s$, we have that 
\begin{equation}
\hat{\eta}_s = \frac{d U_s}{ds} U_s^\dagger = - U_s \frac{d
  U_s^\dagger}{ds} = -\hat{\eta}_s,
\label{eq:eta}
\end{equation} we obtain that 
\begin{equation} \frac{d \hat{H}_s}{ds} = \hat{\eta}_s \hat{H}_s
- \hat{H}_s \hat{\eta}_s = \left[\hat{\eta}_s, \hat{H}_s \right].
\label{eq:flowEquations}
\end{equation} This is the key expression of the SRG method, describing the flow
of the Hamiltonian.  The specific unitary transformation is determined
by the choice of $\hat{\eta}_s$.  Through different choices of
$\hat{\eta}_s$, the SRG evolution can be adapted to the features of a
particular problem.\\

One possibility to solve the flow equations is to choose a basis with
respect to the physical vacuum state, set up the Hamiltonian matrix in
this basis and solve Eq.~(\ref{eq:flowEquations}) as a set of coupled
first-order differential equations. However, since the size of the
problem grows enormously with the number of particles and the size of
the model space, the applicability of this free-space SRG method is
restricted to comparatively small systems.

Instead of performing SRG in free space, the evolution can be done at
finite density, i.e.~directly in the $N$-body system
\cite{kehrein2006flow}. This approach has recently been applied very
successfully in nuclear physics \cite{IMSRG,PhysRevLett.106.222502}
and is called in-medium SRG (IM-SRG). The method allows the evolution
of $3,...,N$-body operators using only two-body machinery, with the
simplifications arising from the use of normal-ordering with respect
to a reference state.

Integrating the flow equations~(\ref{eq:flowEquations}), we face one
of the major challenges of the SRG method, namely the generation of
higher and higher order interaction terms during the flow. With each
evaluation of the commutator, the Hamiltonian gains terms of higher
order, and these induced contributions will in subsequent integration
steps contribute to terms of lower order. In principle, this continues
to infinity.\\ To make the method computationally possible, we have to
close the IM-SRG flow equations, suggesting that we are forced to
truncate the equations to a certain order. We choose to truncate both
$\hat{H}_s$ and $\hat{\eta}_s$ at the two-body level, an approach
which is referred to as IM-SRG(2).  This normal-ordered two-body
approximation seems to be sufficient in many cases and has yielded
excellent results for several nuclei
\cite{PhysRevLett.106.222502,PhysRevLett.109.052501,IMSRG}.\\ The
commutator in the flow equations (\ref{eq:flowEquations}) guarantees
that the IM-SRG wave function $U_s^\dagger|\Phi\rangle$ can be
expanded in terms of linked diagrams only
\cite{shavittbartlett2009,ISI:A1981MN73700014}, which suggests that IM-SRG
is size-extensive. Regarding the quality of the SRG results, it means
that the error introduced by truncating the many-body expansions
scales linearly with the number of particles~$N$. With this
truncation, the generator $\hat{\eta}$ can be written as
%\[
%\hat{\eta} = \sum_{pq} \eta_{pq}^{(1)} \lbrace a_p^\lbrace\dagger\rbrace a_q\rbrace + \frac{1}{4}\sum%\limits_{pqrs}\eta_{pqrs}^{(2)} \lbrace a_p\dagger a_q\dagger a_s a_r \rrb,
%\]
where $\eta_{pq}^{(1)}$ and $ \eta_{pqrs}^{(2)}$ are the one- and
two-body elements, respectively. Making use of the permutation
operator $ \hat{P}_{pq}f(p,q) = f(q,p)$, the IM-SRG(2) flow equations
are given by
Note that for brevity, we skipped all $s$-dependence in the equations.

\subsubsection*{Choice of generator}
To determine the specific unitary transformation, one needs to specify
the generator~$\hat{\eta}$. Through different choices, the SRG flow
can be adapted to the features of a particular problem.\\

Apart from this canonical generator, there exist several other ones in
literature. One of them is White's choice
\cite{White:cond-mat0201346}, which makes numerical approaches much
more efficient.  The problem with Wegner's generator are the widely
varying decaying speeds of the elements, removing first terms with
large energy differences and then subsequently those ones with smaller
energy separations.  That way the flow equations become a stiff set of
coupled differential equations, which often gets numerically
unstable.\\ White takes an alternative approach, which is especially
suited for problems where one is interested in the ground state of a
system. Instead of driving all off-diagonal elements of the
Hamiltonian to zero, he focuses solely on those ones that are
connected to the reference state $|\Phi_0\rangle$, aiming to decouple
the reference state from the remaining Hamiltonian. With a suitable
transformation, the elements get similar decaying speeds, which solves
the problem of stiffness of the flow equations.  The generator is
explicitly constructed the following way
\cite{PhysRevLett.106.222502,White:cond-mat0201346}
\begin{align}
\hat{\eta} &= \sum_{ai} \frac{f_{ai}}{f_a-f_i-v_{aiai}}\lbrace
a_a\dagger a_i\rbrace -\text{hc} \notag \\ & + \sum_{abij}
\frac{v_{abij}}{f_a+f_b-f_i-f_j+A_{abij}}\lbrace a_a\dagger a_b\dagger
a_j a_i\rbrace - \text{hc},
\label{eq:WhiteFull}
\end{align}
with $f_p \equiv f_{pp}$, 'hc' denoting the Hermitian conjugate, and
\[
A_{abij} = v_{abab} + v_{ijij} - v_{aiai} - v_{ajaj} - v_{bibi} -
v_{bjbj}.
\label{eq:White7}
\]
Compared to Wegner's canonical generator, where the final flow
equations involve third order powers of the $f$- and $v$-elements,
these elements contribute only linearly with White's generator, which
results in much better numerical properties.

\subsection{The FCIQMC algorithm}

The recently developed Full Configuration Interaction Quantum Monte
Carlo (FCIQMC) \cite{booth2009,cleland2010,shepherd2012a} promises to
calculate the Full Configuration Interaction (FCI) ground state energy
within a given Hilbert space, and often at a much lower computational
cost than conventional FCI solvers.
%
The FCI energy is given by the equation
\begin{equation}
	E_0^{FCI} = \min{\langle \Psi|\hat H|\Psi\rangle},
	\label{}
\end{equation}
with $|\Psi\rangle\in\mathcal H _{FCI}$. The FCI space $\mathcal H_{FCI}$ is spanned by all
determinants that can be constructed from single-particle states with
an energy lower than the cutoff $\mathbf{k}_c^2/2$.
%
This problem is usually solved by using brute force diagonalizion
techniques like the Lanczos algorithm [mat. comp.], but due to the
large memory footprint of such methods, these are limited to systems
with ``small'' Hilbert spaces. The largest simulations of this kind is
performed with a basis with up to $\sim 10^{11}$(?) determinants [? ?]
which in practice means that only systems with a relatively low number
of particles or small single particle basis can be simulated. The
FCIQMC algorithm does however make it possible threat much larger
Hilbert spaces. One example is the simulations performed by Shepherd
{\it et. al.} \cite{shepherd2012c} where the algorithm is applied on
the three dimensional homogenous electron gas with a bases consisting
of up to $\sim10^{108}$ determinants.

In the following we will briefly introduce the FCIQMC algorithm. 
This is a so called Projector Monte Carlo algorithm where the FCI ground state is found by calculating the projected state
\begin{align}
	\lim_{t\to\infty}e^{-t(\hat H-S)}|\Psi_0\rangle,
	\label{eq:projectedstate}
\end{align}
where the shift $S\in\mathcal R$. The projected state will be
proportional to the FCI ground state $\Psi_0^{FCI}$ if the shift is
equal to $E_0^{FCI}$ and $\expec{\Psi_0|\Psi_0^{FCI}}\neq 0$.  This is
the same approach that is used in other Projector Monte Carlo methods
like Diffusion Monte Carlo or Greens Function Monte Carlo, but the
integration is not performed in the coordinate space but in the
discrete space of Slater determinants. Because of the inherent
anti-symmetric properties of this basis, the infamous sign problem is
not as severe for FCIQMC as for other Projector Monte Carlo methods,
and in principle the exact FCI energies can be calculated. See for
example Refs. \cite{kolodrubetz2013,spencer2012} for a thorough
analysis of the sign problem in FCIQMC.

The large $t$ projected state is found by iterating over the equation
\begin{align}
	|\Psi_0\rangle \gets e^{-\tau(\hat H-S)} |\Psi_0\rangle,
	\label{eqMarkovChain}
\end{align}
where $\tau$ is a small time step. %In this case, $\ket{\Psi_0}$ will become proportional to the FCI ground state in the limit of a large number of iterations.
%
The initial state $|\Psi_0\rangle$ is now expanded on the FCI basis
\begin{align}
	|\Psi_0\rangle = \sum_i C_i |\Phi_i\rangle,
	\label{}
\end{align}
and by approximating the projection operator as its first order Taylor expansion
we can write Eq.~\eqref{eqMarkovChain} in terms of the coefficients $\left\{ C_i \right\}$
\begin{align}
	%\sum_i C_{i}^{(n+1)}\ket{\Phi_i} & \approx (1 - (\hat H - S)\tau) \sum_i C_{i}^{(n)}\ket{\Phi_i}\\
	%&\Downarrow\notag\\
	C_{i} %&= C_{i}^{(n)} - \sum_j(H_{ij}-\partial_{ij}S)\tau C_{j}^{(n)}\notag\\
	%&\gets  C_{i} - (H_{ii}-S)\tau C_{i} - \sum_{j\neq i}H_{ij}\tau C_{j},
	&\gets  C_{i} - \sum_jp_d(i|j) C_{j},
	\label{eq:fciequation}
\end{align}
where we have used $p_d(i|j) = \tau(\hat H_{ij}-S\delta_{ij})$ and
$H_{ij}=\expec{\Phi_i|\hat H|\Phi_j}$.  In the limit of a large number
of iterations, the coefficients $\left\{ C_i \right\}$ will be
proportional to those of the ground state.
%
Note that this approach is exact when the time step $\tau$ is smaller
than $2/(E_M-S)$, where $E_M$ is the largest single particle energy
$E_M=\max(\expec{\Phi_i|\hat H|\Phi_i})$.

A DMC inspired approach is used to find the steady state solution of
Eq. \eqref{eq:fciequation}.  The coefficients $\{C_i\}$ are
represented by a population of $N_W$ walkers $\alpha$ which are
distributed on the determinants $\Phi_{i_\alpha}$ and with a sign
$s_\alpha=\pm$. The amplitude $\{C_i\}$ is then defined to be
\begin{align}
	C_i\propto n_i=\sum_\alpha s_\alpha\delta_{ii_\alpha},
\end{align} 
which is a signed sum where two walkers of opposite signs on the same
determinant will make zero contribution.

It is now possible to define a set of dynamical rules for the walkers
of such a nature that they converge to a distribution
$(n_1,n_2,\dots)\propto (C_1,C_2,\dots)$.  These dynamical rules are
derived from the last term of the right hand side of
Eq. \eqref{eq:fciequation} which describes the rate of which the
coefficients $C_i$ changes during one iteration.

Numerically we do the following. Given a walker on determinant $j$, we
add $N_a=\text{floor}(|p_d(i|j)|)$ walkers on determinant $i$, and one
extra with the probability $N_a-|p_d(i|j)|$. The new walkers have the
same sign as the walker on determinant $j$ is $p_d(i|j)$ is negative
and the opposite sign otherwise.

The sum over all elements $p_d{(i|j)}$ is numerically very expensive
for large FCI bases, and to be able to solve
Eq. \eqref{eq:fciequation} efficiently, Monte Carlo sampling is used
to approximate the sum over the off diagonal terms
\begin{align}
	\sum_j\sum_{i\neq j} p_d{(i|j)} C_{j} \approx \sum_j \sum _{s \in \mathcal S_j} p_d{(s|j)} / p_{gen}(s|j) C_j,\notag\\
	\label{eq:samplingrule}
\end{align}
where $p_{gen}(s|j)$ is the suggestion probability of $s$ from $j$,
and $\mathcal S_j$ consists of $n_j$ elements that are sampled from
$p_{gen}(s|j)$. This reduce the number of operations to $\mathcal
O(N_w)$ each iteration.
%
This equation is only correct when $p_{gen}(s|j)$ is different from
zero for all connected determinants $|\Phi_j\rangle$ and $|\Phi_s\rangle$
\begin{align}
	p_{gen}(s|j)\neq0 \quad\forall\quad H_{js} \neq 0.
\end{align}

In principle, there is no other constraint on $p_{gen}(s|j)$, and as
long as the number of iterations is large enough the simulations will
eventually converge to the correct results, but in practice certain
choices of suggestion probability distributions are expected to lead
to a more efficient sampling.  In our implementation we have used a
$p_{gen}(s|j)$ which is uniform for all $s$ that corresponds to an
interaction $\expec{\Phi_s|\hat H|\Phi_j}$ % \neq 0$ 
where the spin
and momentum are conserved and zero otherwise.

It has been shown that the number of walkers must be above a certain
threshold $N_C$ for the population to converge to the ground state
distribution \cite{spencer2012,bestref?booth?}. This number is known
to be highly system dependent and to increase roughly linearly with
the size of the FCI space \cite{bestref?}.

Cleland {\it et. al.}  (2010) \cite{cleland2010} demonstrated that by
introducing a set of new dynamical rules for the walkers, the
performance of the algorithm can be improved drastically while still
obtaining results with a precision of $10^{-3}$ Ry or better. The
number of walkers that is necessary in such simulations are typically
several orders of magnitude smaller than $N_C$.
%
The improved algorithm is called initiator-FCIQMC or i-FCIQMC, and
does only differ from the original algorithm in how the spawning
probabilites $\{p_d(i|j);i\neq j\}$ are calculated.

First, a so called initiator space is defined. This is a set of
determinants which are assumed to have a relatively large average
population (large ampiludes $|C_i|$). All walkers that live on a
determinant in the initiator space are called initiators, and
initiators follows the same rules as walkers in the basic FCIQMC
algorithm.  The first rule of i-FCIQMC concerns the non-initiators and
states that:
\begin{itemize}
	\item ({\it Spawning rule I}): Non-initiators are only allowed
          to spawn on determinants that are previously populated.
\end{itemize}
With this rule the Hilbert space is in effect reduced to the states in
the initiator space and the single and double excitations of
these. Next, two additional rules are added to improve the energy and
to give the walkers access to the entire FCI space:
\begin{itemize}
			\item ({\it Dynamic enlargement of the
                          initiator space}): We introduce the
                          initiator threshold $N_I$ which is a
                          parameter that is set before a simulations
                          is started.  All determinants $\Phi_i$ with
                          a population $|n_i|\ge N_{I}$ are
                          automatically included in the initiator
                          space for as long the population is above
                          this limit. Here $n_i$ is the signed
                          population of $\Phi_i$. % and $N_i$ is a
                          positive integer that we will refer to as
                          the initiator threshold. The initiator
                          threshold is set before the simulation is
                          started. %Note that with $N_I=1$ i-FCIQMC is
                          identical to FCIQMC.
			\item ({\it Spawning rule II}): If the
                          absolute value of the signed sum
                          $\sum_\alpha s_\alpha$ of non-initiator
                          walkers $\alpha$ that attempts to spawn on
                          an unpopulated determinant is larger than
                          two, then these spawning events are
                          accepted.
\end{itemize}
%
Although the initiator adaption of the algorithm has been shown to
increase the efficiency, it is only correct according to
Eq.~\eqref{eq:fciequation} in the limit of a large number of of
walkers or an initiator threshold $N_I=1$. In this case the initiator
space will grow to include the entire FCI space, and i-FCIQMC will
tend to FCIQMC.
%
This means that the i-FCIQMC energies are systematically improvable
and that the so called initiator error will decay to zero when the
number of walkers is increased.

The initiator spaces are defined using a so called Complete Active
Space (CAS) criterion, which is often used within quantum chemistry
together with methods like many-body pertubation theory or the
configuration interaction method to pick out physically relevant
states of a Hilbert space.
%
Because of the dynamic enlargement of the initiator space, the final
result will be the same regardles of which CAS we choose. But, as
Cleland {\it et. al.} \cite{cleland2010} has shown, an appropriate CAS
might sometimes improve the convergence.

The population of walkers will on average increase if the shift $S$ is larger than $E^{FCI}_0$ and decrease if $S$ is smaller than $E_0^{FCI}$. 
%
We can therefore control the population by varying $S$. We use the formula of Umrigar {\em et al} 
\cite{Umrigar_1993} which has been used for population control in the context of DMC
\begin{align}
	S\to S^{(i)} = S^{(i-1)}- \frac{\xi}\tau \log\left[\frac{N_W^{(i)}}{N_W^{(i-1)}}\right],
	\label{eqvaryshift}
\end{align}
where $\xi$ is a real number of which the optimal value must be found by experiment, and $N_W^{(i)}$ and $S^{(i)}$ are the population and shift at the time $i\tau$. 
%
The shift $S$ that stabilizes the number of walkers must be the ground state energy, and this means 
%
that the above equation provides us with a measure of $E_0^{FCI}$.

We use two different statistical estimators to calculate the energy. The first is the one that we just mentioned, which can be taken as the long time average of the shift 
\begin{align}
	E_0\approx\expec S=(1/N) \sum_{i=1}^N S^{(i)},
	\label{eqvaryshift}
\end{align}
and which we will refer to as the generational estimator of the energy.
%
The second is the projected estimator that is calculated by sampling the projected energy
\begin{align}
	&E_0 = \frac{ \expec{ \Phi_0|\hat H|\Psi_0^{FCI} } }{\expec{ \Phi_0|\Psi_0^{FCI} }} 
	= \frac{ \expec{ \Phi_0|\hat H\sum_iC_i|\Phi_i} }{\expec{ \Phi_0|\sum_iC_i|\Phi_i }} \label{eqProje1},
\end{align}
which is rewritten on the form
\begin{align}
	E_0 \approx &\expec{E_P} = H_{00} + \sum_{i\neq0}H_{0i} \expec{n_i}/\expec{n_0}.
	\label{}
\end{align}

An FCIQMC or i-FCIQMC simulation is usually started with one walker on
the Hartree-Fock (HF) determinant and a shift $S$ that is equal to the
HF energy. Since the HF energy is larger than the FCI energy, the
number of walkers $N_W$ will increase rapidly. To stop the population
growth and stabilize the population around the desired number of
walkers, the shift $S$ is adjusted according to
Eq. \eqref{eqvaryshift}. After a thermalization phase we start
sampling the generational and the projected estimator.  These
estimators are statistical measures, and there will always be a
certain statistical error associated with them. The statistical errors
of both estimators are calculated using Flyvbjerg-Pedersen analysis
[citation], and the energy with the smallest statistical error is
chosen as the final result.

A series of simulations with an increasing number of walkers must be
performed to converge the i-FCIQMC energy. (A study of the convergence
of the initiator error is provided by Sheperd {\it et. al.}
\cite{shepherd2012c}).

@@

\begin{align}
\frac{d E_0}{ds} &= \sum_{ia}\left( \eta_{ia}^{(1)}f_{ai} -
\eta_{ai}^{(1)}f_{ia}\right) +
\frac{1}{2}\sum_{ijab}\eta_{ijab}^{(2)}v_{abij},
\label{eq:flow1}\\
\frac{d f_{pq}}{ds}&= \sum_r \left( \eta_{pr}^{(1)}f_{rq} +
\eta_{qr}^{(1)}f_{rp}\right) + \sum_{ia}\left( 1-\hat{P}_{ia}\right)
\left( \eta_{ia}^{(1)}v_{apiq} - f_{ia}\eta_{apiq}^{(2)} \right)
\notag \\ & +\frac{1}{2} \sum_{aij} \left( 1+\hat{P}_{pq}\right)
\eta_{apij}^{(2)}v_{ijaq} + \frac{1}{2}\sum_{abi}\left(
1+\hat{P}_{pq}\right) \eta_{ipab}^{(2)}v_{abiq},
\label{eq:flow2}
\end{align}

\begin{align}
\frac{d v_{pqrs}}{ds} &= \sum_t \left( 1-\hat{P}_{pq} \right) \left(
\eta_{pt}^{(1)}v_{tqrs}-f_{pt}\eta_{tqrs}^{(2)}\right) -\sum_t \left(
1-\hat{P}_{rs} \right) \left( \eta_{tr}^{(1)} v_{pqts} - f_{tr}
\eta_{pqts}^{(2)}\right) \notag \\ & +\frac{1}{2}\sum_{ab}
\lb\eta_{pqab}^{(2)} v_{abrs} - v_{pqab}\eta_{abrs}^{(2)}\right) -
\frac{1}{2}\sum_{ij} \lb\eta_{pqij}^{(2)} v_{ijrs} -
v_{pqij}\eta_{ijrs}^{(2)}\right) \notag \\ & -\sum_{ia} \left( 1-
\hat{P}_{ia} \right) \left( 1-\hat{P}_{pq}\right) \left(
1-\hat{P}_{rs} \right) \eta_{aqis}^{(2)}v_{ipar}.
\label{eq:flow3}
\end{align}

>>>>

\author{Sean B.~S.~Miller}
\affiliation{Depatment of Physics, Chalmers Technical University, G\"oteborg, Sweden}

@@

@@
